Fitting metabolic models to dissolved oxygen data: the Estuarine BAyesian Single-station Estimation method

### 2023-05-16 manu-draft.qmd
proofing intro and discussion, added acknowledgements and key words

*Key words*:

@@

Methods for estimating metabolism in aquatic environments are numerous and include open-water techniques, bottle-based incubations, ecoystem budgets, oxygen isotopes, and use of inert gases [@kemp2011; @staehr2011]. Based on the variety of assumptions and limitations of each method, intercomparisons of metabolic rates between ecosystem has been challenging. For example, an assessment in Randers Fjord (Denmark) showed agreement in terms of sign and magnitude among four different approaches [@gazeau2005]. However, GPP and ER estimates from incubations were consistently lower than those from open-water methods, consistent with the estuarine respiration synthesis of @hopkinson2005. Even within one technique, there are numerous choices to be made that alter the outcome. For example, in the commonly used 14C technique for measuring primary production, @cloern2014 note that investigators have made divergent choices with regard to filtering mesozooplankton grazers, measuring 14C in dissolved organic carbon, accounting for changing spectral light quality with depth, and choosing the euphotic zone depth, the incubation time, and the incubation location (*in situ* vs. on-deck). These challenges are compounded in estuaries where high temporal and spatial variability have been noted in global syntheses of primary production and respiration. @cloern2014 found that annual phytoplankton primary production can vary 10-fold within an estuary and up to 5-fold from year-to-year. @hopkinson2005 focused on cross-system variations and found benthic and pelagic respiration rates varying by more than factors of 30 and 40, respectively, ranges that are broadly consistent with whole system respiration rates based on the open-water method [@hoellein2013].

Advances in sensor technology for continuous collection of dissolved oxygen time series suggest that the open-water method [@Odum56] could be leveraged for consolidation of approaches. This approach exploits the diel cycle of dissolved oxygen and allow the simultaneous determination of whole-system gross primary production (P, equals net primary production + autotrophic respiration), ecosystem respiration (R, equals heterotrophic + autotrophic respiration), and NEM (equals P-R). A global synthesis of warm-season, open-water metabolism studies [@hoellein2013] was based on a total of 47 estimates, almost all of which were derived from a single study of U.S. estuaries [@caffrey2004]. A similar synthesis was done by @yvon-durocher2012. These studies evaluated the open-water method as applied to continuous monitoring data available for several decades across the National Estuarine Research Reserve System (NERRS). Thus, the approach can be broadly applied where data are available, although it is not without limitations. In particular, tidal advection can violate assumptions of the method that the measurements are from a parcel of water with a continuous metabolic history [@kemp1980]. Striking evidence of this challenge is the large number of physically impossible negative GPP and ER estimates in the open-water analysis of 28 3-year time series by @caffrey2003. The computed rates were negative 23% of the time, on average, and as high as 69% for individual sites. Estimating gas exchange using the open-water method in estuaries has also been challenging. Most approaches estimate gas exchange using empirical formula dependent on wind speed and temperature or held at a constant value [e.g., @caffrey2004; @russell2007; @nagel2009; @Beck15; @Murrell18]. Gas exchange is a first-order process in the estuarine dissolved oxygen budget and, therefore, inaccurate estimation can lead to poorly constrained errors that influence interpretation of results.

A unified model to explain the large variability of metabolic rates within and across estuaries is clearly lacking as a critical tool for understanding metabolic processes. Bayesian techniques have not been fully explored in estuarine applications despite their potential to address the above [although see @ciavatta2008]. Bayesian techniques allow for incorporation of prior knowledge about model parameters and robust estimation of parameter probability distributions [@Holtgrieve10] using highly efficient computational algorithms [@Grace15; @winslow2016]. These techniques are very powerful because they not only estimate GPP and ER as standard open-water methods do, but they retrieve estimates and their uncertainties of gas exchange, GPP, and ER parameters (e.g., the initial slope of the photosynthesis--irradiance curve), which reveal quantitative information about the mechanisms of metabolism and gas exchange. A potentially useful application developed for streams is the BAyesian Single-station Estimation (BASE) method [@Grace15]. BASE has been demonstrated to accurately determine metabolic parameters and their uncertainty using continuous monitoring data that are similar to those available at many estuarine locations. As such, modification of BASE to include model characterizations more appropriate for estuaries, while maintaining the fundamental Bayesian approach, could prove useful.

The goal of this paper is to describe the development and application of a new data assimilation method for simultaneously determining gross primary production, ecosystem respiration, net ecosystem metabolism (NEM), and the gas exchange (D) in estuaries from high-resolution dissolved oxygen time series. We call this method EBASE (Estuarine BAyesian Single-station Estimation), which builds and improves on the BASE method [@Grace15] using an approach more appropriate for estuarine time series. The rationale and changes made to BASE are described in detail, followed by a demonstration of how EBASE can estimate known metabolic parameters with reasonable certainty from a synthetic time series. Lastly, EBASE is compared to metabolic estimates using the Odum open-water method to identify key differences related to both the theoretical and statistical differences associated with each method. These comparisons combine results using real observations and those based on detided inputs to demonstrate effects of tidal advection on metabolic estimates. EBASE is available as an open-source software package created with the R statistical programming language [@Beck23; @RCT23] and the results herein provide practical suggestions for use of the software on novel datasets.

@@

The above analyses demonstrated that EBASE can successfully recover known metabolic parameters from a synthetic time series and provides robust estimates on real data covering a year of observations. The success of these results was predicated on the requirements of EBASE to address several critical needs that existing methods for estimating metabolism do not address for coastal applications. These needs were addressed by 1) using a Bayesian framework that estimates the best fit of the model to the data with posterior probabilities describing the likelihood of the fit, and 2) modifying the underlying mass-balance model for dissolved oxygen with several components more appropriate for estuarine settings. First, the Bayesian framework enables hypothesis testing of how a model can be fit to data by using prior knowledge in the model fitting process, which results in posterior probabilities describing the certainty of the parameter estimates [@Hilborn13]. The advantages of the Bayesian approach have led to widespread adoption over frequentist approaches in many scientific applications, especially in the coastal and marine community [e.g., @myers2001; @borsuk2001; @brown2017], yet these methods have not been rigorously explored for estimating metabolic rates in estuaries. Our results demonstrated that even using uninformed priors, EBASE can reproduce known metabolic parameters and rates with reasonable certainty. However, an exception was the ability of the model to estimate stable values for the $b$ parameter, although the resulting estimate of gas exchange (D) was similar to the known values. This result could be explained by the dominance of the relative processes that manifest signals in the dissolved oxygen time series. That is, estuarine dissolved oxygen at Apalachicola Bay may be dominated by production and respiration, and less so from gas exchange, such that the available data provided insufficient information for EBASE to estimate $b$. This result aligns with others that have suggested gas exchange may be a relatively small component of the metabolic signal in shallow estuaries [@Murrell18]. In such cases, more constrained prior distributions for $b$ could be appropriate or EBASE could be tested in settings with larger contributions of gas exchange expressed by dissolved oxygen.

The second advantage of EBASE that makes the model more appropriate for estuarine settings included several modification to the underlying mass-balance equation for dissolved oxygen (@eq-ebase). As described in @sec-basemodif, EBASE is an estuarine adaptation of BASE (@eq-base) that was developed for stream applications and many assumptions of the underlying model of the latter were inappropriate for the former. Specifically, EBASE was modified to address three primary concerns of using BASE in an estuarine setting. First, the estimation of gross primary production as a function of PAR assumed a linear relationship such that the increase of production per unit increase of PAR was consistent for any level of the latter. The BASE model includes an exponential parameter, $p$, that defines a non-linear relationship between P and PAR (although it should be noted that BASE provides an option to set $p = 1$). @Holtgrieve10 describes a simple linear photosynthesis-irradiance curve as appropriate for systems that are never light-saturated and this assumption likely holds true for estuaries that are commonly more turbid than open-ocean or freshwater locations. Second, EBASE removed the temperature-dependency of respiration and assumed that respiration did not change with water temperature, consistent with applications of the Odum method to other shallow estuaries [@Beck15; @Murrell18]. This modification addresses recent work that suggests respiration in streams may be reduced during the night with lower temperatures as the respiration of labile autochthonous organic matter may contribute to changes in the dissolved oxygen time series [@sadro2014; @schindler2017]. There is currently no justification for this assumption in estuaries. Finally, the gas exchange in EBASE used the parameterization in @Wanninkhof14 that is primarily wind-based and developed for the open-ocean, which is clearly more appropriate than the prior formulation in BASE that includes a temperature and depth dependency for streams. Overall, these changes also reduce the required inputs and many monitoring locations nationwide include the requisite data for applying EBASE (i.e., dissolved oxygen, water temperature, salinity, PAR, and wind speed).

Our results also confirmed that the effects of tidal advection on metabolic estimates can lead to biases in understanding gross primary production and ecosystem respiration. This result has previously been shown using the Odum open water method [@kemp1980; @caffrey2003; @Beck15] and confirmed herein, whereby results using observed time series can produce over-estimates of metabolic rates with more variation than those without tidal influence. Our assessment of the synthetic time series with noise and use of the detided dissolved oyxgen time series from the observed data at Apalachicola Bay to estimate metabolism both demonstrated this bias (Figures [-@fig-synapanoscmp] and [-@fig-apacmpfig]). As a result, detiding dissolved oxygen data is recommended prior to estimating metabolism, regardless of whether EBASE or an alternative method is used. In addition to inaccurate and imprecise metabolic estimates, tidal influences are also most likely observed as negative (or anomalous) values in the production and respiration rates using the Odum method [@caffrey2003]. Tidal effects may be less obvious using EBASE since negative values cannot be obtained based on constraints of the prior distributions, although as noted above, larger and more variable estimates are returned. A potential future approach for developing EBASE could be an explicit formulation of tidal advection in the model using a synoptic tidal height time series in the Bayesian framework. This has the advantage of both detiding and estimating metabolic rates using the same model, whereas using weighted regression prior to EBASE as used herein, requires additional analysis and assumptions. However, field measurements [e.g., control volume approaches, @falter2008; @nidzieko2014] may be required to provide reasonable constraints on the prior distributions for any parameters that quantify advective influences on dissolved oxygen.

EBASE provides several computational advantages over BASE implemented in the R package that was created for the purposes herein [@Beck23]. Most importantly, EBASE was developed to allow different model optimization lengths, whereas the BASE method was developed to estimate results for a single day. Allowing EBASE to incorporate more than a day of observations in the optimization can likely produce more accurate and precise estimates for the metabolic parameters. Physical and biological characteristics of an ecosystem that affect the metabolic rates are likely acting at time scales longer than a day, e.g., wind patterns that can affect gas exchange as quantified by the $b$ parameter may persist for several days. Exposing the model to additional observations may produce more stable results as the parameters for the best fit are estimated by the Bayesian routine as a function of the data that are evaluated each optimization period. However, we cannot provide a precise recommendation on the number of days to use for the optimization period as the exact length of time that ecosystem characteristics can affect metabolic rates may not be known and likely varies by location. Our example using the synthetic time series from Apalachicola Bay suggests that results were similar for the 7 and 30 day optimization period. As such, using one week of data for each optimization period is likely appropriate for this location. Using one day is not recommended given the issues stated above (i.e., unstable results), whereas an optimization period longer than a month may miss important events in the time series that can affect the metabolic estimates.

The EBASE R package can also accommodate missing observations in the input data. This allows the estimation of metabolic rates for time series that cover relatively long periods of time (e.g., more than year), when data gaps are likely to occur for several reasons (e.g., equipment malfunction, routine maintenance, etc.). The R package accommodates data gaps using linear interpolation for any required parameter used by EBASE. This allows the Bayesian routine to provide continuous estimates across the time series. The results are then post-processed such that any metabolic estimates that exceed a threshold for interpolated observations are removed from the output. The analyst can determine the appropriate length of time to use for excluding results based on interpolated values. The default setting for EBASE is 12 hours, where results for an entire optimization period are removed if any interpolated period exceeded 12 hours within that period. Although this does not apply to the synthetic time series, some of the estimates using the 2021 observed data for Apalachicola Bay were excluded from the analysis. This feature allows EBASE to be applied more easily to other data setes because an analyst does not need to manually screen data that include missing values.

While the EBASE theory and software provide advantages for metabolic estimates, the approach is not without limitations. Like most models, explicit choices are needed to apply EBASE to time series data. First, the use of unconstrained prior distributions was shown to reproduce known metabolic parameters with reasonable certainty, although the $b$ parameter was still relatively unstable and unconstrained priors may not provide useful estimates in estuarine settings outside of the location evaluated herein. This may require further constraining the priors, although doing so may produce biased results in the absence of additional data to justify a constraint. Alternative metabolic models could be used for informing priors, such as using the Odum method to inform the prior for the reareation coefficient, $K$, for the BASE method. Initial application of BASE to estuarine data demonstrated that unstable and unrealistic metabolic rates were returned with uninformed priors unless the $K$ parameter was constrained. A similar approach could be applied to create more accurate estimates from EBASE depending on location. An additional choice for using EBASE is the model optimization period, although some guidance is provided above.

@@

The development of EBASE represents a new approach for estimating metabolic rates in estuarine settings that leverages existing methods by adapting theoretical components of the underlying mass-balance model using a Bayesian framework. Our results have demonstrated that unknown metabolic parameters can be estimated with reasonable certainty using uninformed prior distributions and EBASE could be a valuable approach to improve the understanding of critical ecosystem processes as the method can be readily applied to other settings. A notable limitation of EBASE was an inability to precisely recover the $b$ parameter from a synthetic time series, although this result may be explained by a minimal contribution of gas exchange as expressed in the synthetic dissolved oxygen time series. Effects of tidal advection on dissolved oxygen measurements were also observed, demonstrating a violation of the assumption that the measured data represent a water sample having a continuous metabolic history. Further development of EBASE to accommodate a tidal signal, or exploring alternative methods for detiding dissolved oxygen data, are promising avenues of future research to more accurately quantify metabolic parameters in estuaries. Overall, EBASE represents a potentially powerful tool that removes many of the assumptions of existing methods by allowing unknown metabolic parameters to be estimated by the data and can also incorporate existing knowledge for further exploration of estuarine ecosystem properties.

@@

>>>>

*Key words*: Bayesian, dissolved oxygen, ecosystem metabolism, estuary, gas exchange, open source

@@

Methods for estimating metabolism in aquatic environments are numerous and include open-water techniques, bottle-based incubations, ecoystem budgets, oxygen isotopes, and use of inert gases [@kemp2011; @staehr2011]. Based on the variety of assumptions and limitations of each method, comparisons of metabolic rates within and between ecosystems hav been challenging. For example, an assessment in Randers Fjord (Denmark) showed agreement in both the sign and magnitude among four different approaches [@gazeau2005]. However, gross primary production (P) and ecosystem respiration (R) estimates from incubations were consistently lower than those from open-water methods, consistent with the estuarine respiration synthesis of @hopkinson2005. Even within one technique, there are numerous choices to be made that alter the outcome. For example, in the commonly used ${}^{14}$C technique for measuring primary production, @cloern2014 noted that investigators have made divergent choices with regard to filtering mesozooplankton grazers, measuring ${}^{14}$C in dissolved organic carbon, accounting for changing spectral light quality with depth, and choosing the euphotic zone depth, the incubation time, and the incubation location (*in situ* vs. on-deck). These challenges are further compounded in estuaries where high temporal and spatial variability have been noted in global syntheses of P and R. @cloern2014 found that annual phytoplankton primary production can vary 10-fold within an estuary and up to 5-fold from year-to-year. @hopkinson2005 focused on cross-system variations and found benthic and pelagic respiration rates varying by more than factors of 30 and 40, respectively, ranges that are broadly consistent with whole system respiration rates based on the open-water method [@hoellein2013].

Advances in sensor technology for continuous collection of dissolved oxygen time series suggest that the open-water method [@Odum56] could be leveraged for applications across multiple locations. This approach exploits the diel cycle of dissolved oxygen and allows the simultaneous determination of whole-system gross primary production (net primary production + autotrophic respiration), ecosystem respiration (equals heterotrophic + autotrophic respiration), and NEM (equals P-R). A global synthesis of warm-season, open-water metabolism studies [@hoellein2013] was based on a total of 47 estimates, almost all of which were derived from a single study of U.S. estuaries [@caffrey2004]. A similar synthesis was done by @yvon-durocher2012. These studies evaluated the open-water method as applied to continuous monitoring data available for several decades across the National Estuarine Research Reserve System (NERRS). Thus, the approach can be broadly applied where data are available, although it is not without limitations. In particular, tidal advection can violate assumptions of the method that the measurements are from the same parcel of water with a continuous metabolic history [@kemp1980]. Striking evidence of this challenge is the large number of physically impossible negative P and R estimates in the open-water analysis of 28 3-year time series by @caffrey2003. The computed rates were negative 23% of the time, on average, and as high as 69% for individual sites. Estimating gas exchange using the open-water method in estuaries has also been challenging. Most approaches estimate gas exchange using empirical formula dependent on wind speed and temperature or held at a constant value [e.g., @caffrey2004; @russell2007; @nagel2009; @Beck15; @Murrell18]. Gas exchange is a first-order process in the estuarine dissolved oxygen budget and, therefore, inaccurate estimations based on empirical formula or fixed values can lead to poorly constrained errors that influence interpretation of results.

A unified model to explain the large variability of metabolic rates within and across estuaries is clearly lacking as a critical tool for understanding metabolic processes. Bayesian techniques have not been fully explored in estuarine applications despite their potential to address the above [although see @ciavatta2008]. Bayesian techniques allow for incorporation of prior knowledge about model parameters and robust estimation of parameter probability distributions [@Holtgrieve10] using highly efficient computational algorithms [@Grace15; @winslow2016]. These techniques are very powerful because they not only estimate P and R as standard open-water methods do, but they retrieve estimates and their uncertainties of gas exchange, P, and R parameters (e.g., the initial slope of the photosynthesis--irradiance curve), which reveals quantitative information about the mechanisms of metabolism and gas exchange. A potentially useful application developed for streams is the BAyesian Single-station Estimation (BASE) method [@Grace15]. BASE has been demonstrated to accurately determine metabolic parameters and their uncertainty using continuous monitoring data that are similar to those available at many estuarine locations. As such, modification of BASE to include model characterizations more appropriate for estuaries, while maintaining the fundamental Bayesian approach, could prove useful.

The goal of this paper is to describe the development and application of a new Bayesian method for simultaneously determining gross primary production, ecosystem respiration, net ecosystem metabolism, and the gas exchange (D) in estuaries from high-resolution dissolved oxygen time series. We call this method EBASE (Estuarine BAyesian Single-station Estimation), which builds and improves on the BASE method [@Grace15] using an approach more appropriate for estuarine time series. The rationale and changes made to BASE are described in detail, followed by a demonstration of how EBASE can estimate known metabolic parameters with reasonable certainty from a synthetic time series. Lastly, EBASE is compared to metabolic estimates using the Odum open-water method to identify key differences related to both the theoretical and statistical differences associated with each method. These comparisons combine results using real observations and those based on detided inputs to demonstrate effects of tidal advection on metabolic estimates. EBASE is available as an open-source software package created with the R statistical programming language [@Beck23; @RCT23] and the results herein provide practical suggestions for use of the software on novel datasets.

@@

The above analyses demonstrated that EBASE can successfully recover known metabolic parameters from a synthetic time series and provides robust estimates on real data covering a year of observations. The success of these results was predicated on the requirements of EBASE to address several critical needs that existing methods for estimating metabolism do not address for coastal applications. These needs were addressed by 1) using a Bayesian framework that estimates the best fit of the model to the data with posterior probabilities describing the likelihood of the fit, and 2) modifying the underlying mass-balance model for dissolved oxygen with several components more appropriate for estuarine settings. First, the Bayesian framework enables hypothesis testing of how a model can be fit to data by using prior knowledge in the model fitting process, which results in posterior probabilities describing the certainty of the parameter estimates [@Hilborn13]. The advantages of the Bayesian approach have led to widespread adoption over frequentist approaches in many scientific applications, especially in the coastal and marine community [e.g., @myers2001; @borsuk2001; @brown2017], yet these methods have not been rigorously explored for estimating metabolic rates in estuaries. Our results demonstrated that even using uninformed priors, EBASE can reproduce known metabolic parameters and rates with reasonable certainty. However, an exception was the ability of the model to estimate stable values for the $b$ parameter, although the resulting estimate of gas exchange was similar to the known values. This result could be explained by the dominance of the relative processes that manifest signals in the dissolved oxygen time series. That is, estuarine dissolved oxygen at Apalachicola Bay may be dominated by production and respiration, and less so from gas exchange, such that the available data provided insufficient information for EBASE to estimate $b$. This result aligns with others that have suggested gas exchange may be a relatively small component of the metabolic signal in shallow estuaries of the Gulf of Mexico [@Murrell18]. In such cases, more constrained prior distributions for $b$ could be appropriate or EBASE could be tested in settings with expected larger contributions of gas exchange expressed by dissolved oxygen.

The second advantage of EBASE that makes the model more appropriate for estuarine settings included several modification to the underlying mass-balance equation for dissolved oxygen (@eq-ebase). As described in @sec-basemodif, EBASE is an estuarine adaptation of BASE (@eq-base) that was developed for stream applications and many assumptions of the underlying model of the latter were inappropriate for the former. Specifically, EBASE was modified to address three primary concerns of using BASE in an estuarine setting. First, the estimation of gross primary production as a function of PAR assumed a linear relationship such that the increase of production per unit increase of PAR was consistent for any level of the latter. The BASE model includes an exponential parameter, $p$, that defines a non-linear relationship between P and PAR (although it should be noted that BASE provides an option to set $p = 1$). @Holtgrieve10 describes a simple linear photosynthesis-irradiance curve as appropriate for systems that are never light-saturated and this assumption likely holds true for estuaries that are commonly more turbid than open-ocean or freshwater locations. Second, EBASE removed the temperature-dependency of respiration and assumed that respiration did not change with water temperature, consistent with applications of the Odum method to other shallow estuaries [@Beck15; @Murrell18]. This modification addresses recent work that suggests respiration in streams may be reduced during the night with lower temperatures as the respiration of labile autochthonous organic matter may contribute to changes in the dissolved oxygen time series [@sadro2014; @schindler2017]. There is currently no justification for this assumption in estuaries. Finally, the gas exchange in EBASE used the parameterization in @Wanninkhof14 that is primarily wind-based and developed for the open-ocean, which is clearly more appropriate than the prior formulation in BASE that includes a temperature and depth dependency for streams. Overall, these changes also reduced the required inputs and many monitoring locations nationwide include the requisite data for applying EBASE (i.e., dissolved oxygen, water temperature, salinity, PAR, and wind speed, as collected by the NERRS).

Our results also confirmed that the effects of tidal advection on metabolic estimates can lead to biases in understanding gross primary production and ecosystem respiration. This result has previously been shown using the Odum open water method [@kemp1980; @caffrey2003; @Beck15] and confirmed herein, whereby results using observed time series can produce over-estimates of metabolic rates with more variation than those without tidal influence. Our assessment of the synthetic time series with noise and use of the detided dissolved oyxgen time series from the observed data at Apalachicola Bay to estimate metabolism both demonstrated this bias (Figures [-@fig-synapanoscmp] and [-@fig-apacmpfig]). As a result, detiding dissolved oxygen data is recommended prior to estimating metabolism, regardless of whether EBASE or an alternative method is used. In addition to inaccurate and imprecise metabolic estimates, tidal influences are also most likely observed as negative (or anomalous) values in the production and respiration rates using the Odum method [@caffrey2003]. Tidal effects may be less obvious using EBASE since negative values cannot be obtained based on constraints of the prior distributions, although as noted above, larger and more variable estimates are returned. A potential future approach for developing EBASE could be an explicit formulation of tidal advection in the model using a synoptic tidal height time series in the Bayesian framework. This has the advantage of simultaneously detiding and estimating metabolic rates using the same model, whereas using weighted regression prior to EBASE, as used herein, requires additional analysis and assumptions. However, field measurements [e.g., control volume approaches, @falter2008; @nidzieko2014] may be required to provide reasonable constraints on the prior distributions for any parameters that quantify advective influences on dissolved oxygen.

EBASE provides several computational advantages over BASE implemented in the R package that was created for the purposes herein [@Beck23]. Most importantly, EBASE was developed to allow different model optimization lengths, whereas the BASE method was developed to estimate results for a single day. Allowing EBASE to incorporate more than a day of observations in the optimization can likely produce more accurate and precise estimates for the metabolic parameters. Physical and biological characteristics of an ecosystem that affect the metabolic rates are likely acting at time scales longer than a day, e.g., wind patterns that can affect gas exchange as quantified by the $b$ parameter may persist for several days. Exposing the model to additional observations may produce more stable results as the parameters for the best fit are estimated by the Bayesian routine as a function of the data that are evaluated each optimization period. However, we cannot provide a precise recommendation on the number of days to use for the optimization period as the exact length of time that ecosystem characteristics can affect metabolic rates may not be known and likely varies by location. Our example using the synthetic time series from Apalachicola Bay suggests that results were similar for the 7 and 30 day optimization period. As such, using one week of data for each optimization period is likely appropriate for this location. For other locations, using one day is not recommended given the issues stated above (i.e., unstable results), whereas an optimization period longer than a month may miss important events in the time series that can affect the metabolic estimates and will likely produce parameters that are overly generalized.

The EBASE R package can also accommodate missing observations in the input data. This allows the estimation of metabolic rates for time series that cover relatively long periods of time (e.g., more than year), when data gaps are likely to occur for several reasons (e.g., equipment malfunction, routine maintenance, etc.). The R package accommodates data gaps using linear interpolation for any required input used by EBASE. This allows the Bayesian routine to provide continuous estimates across the time series. The results are then automatically post-processed to remove any metabolic estimates that exceed a threshold for the number of interpolated observations. The analyst can determine the appropriate length of time to use for excluding results based, although the default setting for EBASE is 12 hours, where results for an entire optimization period are removed if any interpolated period exceeded 12 hours within that period. Although this does not apply to the synthetic time series, some of the estimates using the 2021 observed data for Apalachicola Bay were excluded from the analysis. This feature allows EBASE to be applied more easily to other data sets because an analyst does not need to manually screen data that include missing values prior to analysis.

While the EBASE theory and software provide advantages for metabolic estimates, the approach is not without limitations. Like most models, explicit choices are needed to apply EBASE to time series data. First, the use of unconstrained prior distributions was shown to reproduce known metabolic parameters with reasonable certainty, although the $b$ parameter was still relatively unstable and unconstrained priors may not provide useful estimates in estuarine settings outside of the location evaluated herein. This may require further constraining the priors, although doing so may produce biased results in the absence of additional data to justify a constraint. Field-based measurements or alternative metabolic models could be used for informing priors. As an example, initial application of BASE to estuarine data prior to developing EBASE demonstrated that unstable and unrealistic metabolic rates were returned with uninformed priors. The Odum method was used on the same data to develop an appropriate expectation of the range of values for the reaeration coefficient, $K$ , which was then used to inform the prior distribution using the BASE method. More stable and realistic metabolic estimates were then returned by BASE, demonstrating the value of the Bayesian approach that can incorporate prior knowledge. A similar approach could be applied to create more accurate estimates from EBASE depending on location. An additional choice for using EBASE is the model optimization period, although some guidance is provided above.

@@

The development of EBASE represents a new approach for estimating metabolic rates in estuarine settings that leverages existing methods by adapting theoretical components of the underlying mass-balance model using a Bayesian framework. Our results have demonstrated that unknown metabolic parameters can be estimated with reasonable certainty using uninformed prior distributions and EBASE could be a valuable approach to improve the understanding of critical ecosystem processes as the method can be readily applied to other settings. A notable limitation of EBASE was an inability to precisely recover the $b$ parameter from a synthetic time series, although this result may be explained by a minimal contribution of gas exchange as expressed in the dissolved oxygen time series. Effects of tidal advection on dissolved oxygen measurements were also observed, demonstrating a violation of the assumption that the measured data represent a water sample having a continuous metabolic history. Further development of EBASE to accommodate a tidal signal, or exploring alternative methods for detiding dissolved oxygen data, are promising avenues of future research to more accurately quantify metabolic parameters in estuaries. Overall, EBASE represents a potentially powerful tool that removes many of the assumptions of existing methods by allowing unknown metabolic parameters to be estimated by the data and can also incorporate existing knowledge for further exploration of estuarine ecosystem properties.

@@

We thank Jason Garwood and Ethan Borque of the Apalachicola National Estuarine Research Reserve for assisting with the deployment and processing of data used herein. We also thank Wade McGillis for assisting in this effort. We thank Darren Giling for initial conversations on applying BASE to estuarine data. This research was funded by award 1924559 from the National Science Foundation to co-PIs Ray Najjar, Maria Herrmann, and Kathleen Hill.
